2.6: Inference for Regression

### 2019-10-08 static/slides/12-slides.Rmd
Update slides 12

---
title: "2.6: Inferential Statistics I"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

@@

>>>>

---
title: "2.6: Inference for Regression"
subtitle: "ECON 480 · Econometrics · Fall 2019"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i> safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/metricsf19"><i class="fa fa-github fa-fw"></i> ryansafner/metricsf19</a><br> <a href="https://metricsF19.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i> metricsF19.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    df_print: paged
    #seal: false
    css: [custom.css, custom-fonts.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

@@

# Classical Statistical Inference: Critical Values of Test Statistic

- .whisper[Test statistic `\\((\delta)\\)`]: measures **how far what we observed in our sample $(\hat{\beta_1})$ is from what we would expect if the null hypothesis were true $(\beta_1=0)$**
    - Calculated from a sampling distribution of the estimator (i.e. $\hat{\beta_1})$
    - In econometrics, we use $t$-distributions which have $n-k-1$ degrees of freedom<sup>.red</sup>

- .whisper[Rejection region]: if the test statistic reaches a .onfire["critical value"] of $\delta$, then we **reject** the null hypothesis

.footnote[<sup>.red[1]</sup> Again, see today's [class notes](/class/12-class) for more on the t-distribution. `\\(k\\)` is the number of independent variables our model has, in this case, with just one `\\(X\\)`, `\\(k=1\\)`. We use two degrees of freedom to calculate `\\(\hat{\beta_0}\\)` and `\\(\hat{\beta_1}\\)`, hence we have `\\(n-2\\)` df.]

# Simulating the Sampling Distribution with *infer*

# Imagine a Null World, where $H_0$ is True

.center[
![:scale 70%](https://www.dropbox.com/s/x7b5azt5algjodg/parallelworld.jpeg?raw=1)

Our world, and a world where $\beta_1=0$ by assumption.
]

# Comparing the Worlds I

- From that null world where $H_0: \, \beta_1=0$ is true, we **simulate** another sample and calculate OLS estimators again

# Our Sample

```{r, echo=F}
tidy(school_reg)
```
]

# Another Sample

```{r, echo = F}
library(infer)
CASchool %>%
  specify(testscr ~ str) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1, type = "permute") %>%
  lm(testscr ~ str, data = .) %>%
  tidy()
```
]

# Comparing the Worlds II

- From that null world where $H_0: \, \beta_1=0$ is true, let's **simulate 1,000** samples and calculate slope $(\hat{\beta_1})$ for each

```{r}
CASchool %>%
  specify(testscr ~ str) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "slope")%>%
  rename(sample = replicate,
         slope = stat)
```

# Prepping the *infer*  Pipeline

- Before I show you how to do this, let's first save our estimated slope from our *actual* sample
    - We'll want this later!

```{r}
# save as obs_slope
sample_slope <- school_reg_tidy %>% # this is the regression tidied with broom
  filter(term=="str") %>%
  pull(estimate)

# confirm what it is
sample_slope
```

---
# The *infer*  Pipeline: Specify

# The *infer*  Pipeline: Specify

.left-code[
### Specify
```{r, echo=T, eval=F}
data %>%
  specify(y ~ x)
```
]

.right-plot[
- Take our data and pipe it into the `specify()` function, which is essentially a `lm()` function for regression (for our purposes)

```{r, echo=T, eval=F}
CASchool %>%
  specify(testscr ~ str)%>%
  head(.,3)
```

```{r}
CASchool %>%
  specify(testscr ~ str)%>%
  head(.,3)
```
]

- Note nothing happens yet

# The *infer*  Pipeline: Hypothesize

# The *infer*  Pipeline: Hypothesize

.left-code[
### Specify
### Hypothesize
```{r, echo=T, eval=F}
%>% hypothesize(null = "independence")
```
]

.right-plot[
- Describe what the null hypothesis is here
- In `infer`'s language, we are hypothesizing that `str` and `testscr` are `independent` $(\beta_1=0)$<sup>.red[1]</sup>

```{r, echo=T, eval=F}
CASchool %>%
  specify(testscr ~ str) %>%
  hypothesize(null = "independence")
```

```{r}
CASchool %>%
  specify(testscr ~ str) %>%
  hypothesize(null = "independence") %>%
  head(.,3)
```
]

- Note nothing happens yet

.footnote[<sup>.red[1]</sup> `type` can be either `point` (for specific point estimates for a single variable, such as a sample mean, `\\((\bar{x})\\)`, or `independence` (for hypotheses about two samples or a relationship between variables). See more [here](https://moderndive.netlify.com/9-hypothesis-testing.html).]
---

# The *infer*  Pipeline: Generate I

# The *infer*  Pipeline: Generate I

.left-code[
### Specify
### Hypothesize
### Generate
```{r, echo=T, eval=F}
%>% generate(reps = n, type = "permute")
```
]

.right-plot[
- Now the magic starts, as we run a number of simulated samples<sup>.red[1]</sup>
- Set the number of `reps` and set the `type` equal to `"permute"`

```{r, echo=T, eval=F}
i %>%
  generate(reps = 1000, type = "permute")
```

.footnote[<sup>.red[1]</sup> Note for spacing on the slide, I saved the previous code as `i` and pipe it into the remainder.]

# The *infer*  Pipeline: Generate II

.left-code[
### Specify
### Hypothesize
### Generate
```{r, echo=T, eval=F}
%>% generate(reps = n, type = "permute")
```
]

.right-plot[

```{r}
i<-CASchool %>%
  specify(testscr ~ str) %>%
  hypothesize(null = "independence")
i %>%
  generate(reps = 1000, type = "permute")
```
]

# The *infer*  Pipeline: Generate III

.left-code[
### Specify
### Hypothesize
### Generate
```{r, echo=T, eval=F}
%>% generate(reps = n, type = "permute")
```
]

.right-plot[

- There are two types of simulations we can run:<sup>.red[1]</sup>
- `"bootstrap"` takes a random draw of our *existing* sample's observations (of the same number of observations) *with replacement*
  - this approximates a sampling distribution
- `"permute"` is a `bootstrap` *without* replacement
]

.footnote[<sup>.red[1]</sup> You can do either of these in base R with `sample()`, which has 3 arguments: a vector to sample from, `size` (number of obs), and `replace` equal to `TRUE` or `FALSE`. See more for `infer`  [here](https://moderndive.netlify.com/9-hypothesis-testing.html).]

# The *infer*  Pipeline: Calculate I

# The *infer*  Pipeline: Calculate I

.left-code[
### Specify
### Hypothesize
### Generate
### Calculate
```{r, echo=T, eval=F}
%>% calculate(stat = "")
```
]

.right-plot[
- We `calculate` sample statistics for each of the 1,000 `replicate` samples

- In our case, calculate the slope, $(\hat{\beta_1})$ for each `replicate`

```{r, echo=T, eval=F}
i %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "slope")
```

- Other `stat`s for calculation: `"mean"`, `"median"`, `"prop"`, `"diff in means"`, `"diff in props"`, etc. (see [package information](https://infer.netlify.com/))
]

# The *infer*  Pipeline: Calculate II

.left-code[
### Specify
### Hypothesize
### Generate
### Calculate
```{r, echo=T, eval=F}
%>% calculate(stat = "")
```
]

.right-plot[
```{r, echo=F}
i %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "slope")
```
]

# The *infer*  Pipeline: Get p Value

.left-code[
### Specify
### Hypothesize
### Generate
### Calculate
### Get p Value
```{r, echo=T, eval=F}
%>% get_p_value(obs stat = "",
                direction = "both")
```
]

.right-plot[

- We can calculate the .shout[`\\(p\\)`-value]
    - the probability of seeing a value at least as large as our `sample_slope` (`r round(sample_slope, 3)`) in our simulated null distribution

- onfire[Two-sided alternative] $H_a: \beta_1 \neq 0$, we double the raw $p$-value

```{r, echo=F}
simulations <- i %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "slope")
```

```{r, echo=T}
simulations %>%
  get_p_value(obs_stat = sample_slope,
              direction = "both")
```
]

.footnote[<sup>.red[+]</sup> Note here I saved the results of our previous code as `simulations` for spacing.]

# The *infer*  Pipeline: Get Confidence Interval

.left-code[
### Specify
### Hypothesize
### Generate
### Calculate
### Get confidence interval
```{r, echo=T, eval=F}
%>% get_ci(
  level = 0.95,
  type = "se",
  point_estimate = "")
```
]

.right-plot[

- We can calculate the .shout[confidence interval] for $\beta_1$ from our `sample_slope` $\hat{\beta_1}$ of `r round(sample_slope, 3)`
    - $CI_{0.95}= \hat{\beta_1} \pm 2 \times se(\hat{\beta_1})$
    - Specify confidence level $(1-\alpha)$, often $0.95$
    - Supply our original estimate for `point_estimate`

```{r, echo=T}
simulations %>%
  get_confidence_interval(level = 0.95,
                          type = "se",
                          point_estimate = sample_slope)
```
]

# The *infer*  Pipeline: Visualize I

# The *infer*  Pipeline: Visualize I

.left-code[
### Specify
### Hypothesize
### Generate
### Calculate
### Visualize
```{r, echo=T, eval=F}
%>% visualize()
```
]

.right-plot[

- Make a histogram of our null distribution of $\beta_1$
    - Note it is centered at $\beta_1=0$ because that's $H_0$!

```{r, echo=T, fig.height=4, fig.retina=3, fig.align="center"}
simulations %>%
 visualize()
```
]

# The *infer*  Pipeline: Visualize II

.left-code[
### Specify
### Hypothesize
### Generate
### Calculate
### Visualize
```{r, echo=T, eval=F}
%>% visualize()
```
]

.right-plot[

- Add our `sample_slope` to show our finding on the null distr.

```{r, echo=T, fig.height=4, fig.retina=3, fig.align="center"}
simulations %>%
 visualize(obs_stat = sample_slope)
```
]

# The *infer*  Pipeline: Visualize p-value

.left-code[
### Specify
### Hypothesize
### Generate
### Calculate
### Visualize
```{r, echo=T, eval=F}
%>% visualize()+
  shade_p_value()
```
]

.right-plot[

- Add `shade_p_value` to see what $p$ is

```{r, echo=T, fig.height=4, fig.retina=3, fig.align="center"}
simulations %>%
 visualize(obs_stat = sample_slope)+
  shade_p_value(obs_stat = sample_slope,
                direction = "two_sided")
```
]

# The *infer*  Pipeline: Visualize Confidence Intervals

.left-code[
### Specify
### Hypothesize
### Generate
### Calculate
### Visualize
```{r, echo=T, eval=F}
%>% visualize()+
  shade_ci()
```
]

.right-plot[

- To shade confidence interval, we first need a vector of what they are
  - I've saved the outputted `tibble` of them from 4 slides ago as `ci_values`

```{r}
ci_values<-simulations %>%
  get_confidence_interval(level = 0.95,
                          type = "se",
                          point_estimate = sample_slope)
```

```{r, echo=T, fig.height=4, fig.retina=3, fig.align="center"}
simulations %>%
 visualize(obs_stat = sample_slope)+
  shade_confidence_interval(ci_values)
```
]

# The *infer*  Pipeline: Visualize is a Wrapper of ggplot

- `infer`'s `visualize()` function is just a wrapper function for `ggplot()`
  - you can take your `simulations` `tibble` and just `ggplot` a normal histogram

.pull-left[
```{r visualize-gg, echo=T, eval=F}
simulations %>%
  ggplot(data = .)+
  aes(x = stat)+
  geom_histogram(color="white", fill="indianred")+
  geom_vline(xintercept = sample_slope,
             color = "blue",
             size = 2,
             linetype = "dashed")+
  labs(x = expression(paste("Distribution of ", hat(beta[1]), " under ", H[0], " that ", beta[1]==0)),
       y = "Samples")+
    theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```
]

.pull-right[
```{r, ref.label="visualize-gg", fig.height=5, fig.retina=3, fig.align="center"}

# What R Does: Classical Statistical Inference I

- R does things the old-fashioned way, using a *theoretical* null distribution instead of *simulation*

- A .shout[`\\(t\\)`-distribution] with $n-k-1$ df<sup>1</sup>

- Calculate a $t$-statistic for $\hat{\beta_1}$:

$$\text{test statistic} = \frac{\text{estimate} - \text{null hypothesis}}{\text{standard error of estimate}}$$

```{r, fig.retina=3, fig.align="center"}
t_ex<-ggplot(data = tibble(x=-5:5))+
  aes(x = x)+
  stat_function(fun = dt, args=list(df = 418), size=2, color="blue")+
  #stat_function(fun=dt, geom="area", fill="blue", xlim=c(-5,4.75), alpha=0.5)+
  #stat_function(fun=dt, geom="area", fill="blue", xlim=c(4.75,5), alpha=0.5)+
  geom_vline(xintercept = c(-4.75, 4.75), size = 1, color = "red", linetype = "dashed")+
  geom_rect(xmin=-5,xmax=-4.75, ymin=0, ymax=0.4, fill="red", alpha=0.3)+
  geom_rect(xmin=4.75,xmax=5, ymin=0, ymax=0.4, fill="red", alpha=0.3)+
  labs(x = "t",
       y = "Probability")+
  scale_x_continuous(breaks = seq(-5,5,1),
                     lim = c(-5,5),
                     expand = c(0,0))+
  scale_y_continuous(breaks = seq(0,0.4,0.1),
                     lim = c(0,0.4),
                     expand = c(0,0))+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
t_ex
```

.footnote[<sup>1</sup> $k$ is the number of $X$ variables.]

# What R Does: Classical Statistical Inference II

$$\text{test statistic} = \frac{\text{estimate} - \text{null hypothesis}}{\text{standard error of estimate}}$$

- $t$ has the same interpretation as $Z$, number of std. dev. away from the distribution's center<sup>1</sup>

- Compares to a critical value of $t^*$ (determined by $\alpha$ & $n-k-1$)
    - For 95% confidence, $\alpha=0.05$, $t^* \approx 2$<sup><span class="red">2</span></sup>
]

```{r, fig.retina=3, fig.align="center"}
t_ex
```

.footnote[<sup>1</sup> Think of our simulated distribution, the center was 0.

<sup>2</sup> The 68-95-99.7% empirical rule!]

# What R Does: Classical Statistical Inference III

$$\begin{align*}
t &= \frac{\hat{\beta_1}-\beta_{1,0}}{se(\hat{\beta_1})}\\
t &= \frac{-2.28-0}{0.48}\\
t &= -4.75\\ \end{align*}$$

- Our sample slope is 4.75 standard deviations below the mean under $H_0$

- $p$-value: prob. of a test statistic at least as large (in magnitude) as ours if the null hypothesis were true<sup>1</sup>
  - $p$-value is **2-sided** for $H_a: \beta_1 \neq 0$
]

```{r, fig.retina=3, fig.align="center"}
t_ex
```

.footnote[<sup>1</sup> Think of our simulated distribution, the center was 0.]

# 1-Sided vs. 2-Sided p-values I

$H_a: \beta_1<0$

$p$-value: $Prob(t < t_i)$

```{r, fig.retina=3, fig.height=5, fig.align="center"}
ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  stat_function(fun=dnorm, xlim=c(-4,-2), geom="area", fill="blue", alpha=0.5)+
  labs(x = "t",
       y = "Probability")+
  scale_x_continuous(breaks=c(-2),
                      label=c(expression(-t[i])))+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)

]
.pull-right[

$H_a: \beta_1>0$

$p$-value: $Prob(t > t_i)$

```{r, fig.retina=3, fig.height=5, fig.align="center"}
ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  stat_function(fun=dnorm, xlim=c(2,4), geom="area", fill="blue", alpha=0.5)+
  labs(x = "t",
       y = "Probability")+
  scale_x_continuous(breaks=c(2),
                      label=c(expression(t[i])))+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

# 1-Sided vs. 2-Sided p-values I

$H_a: \beta_1 \neq 0$

$p$-value: $2 \times Prob(t > |t_i|)$

```{r, fig.retina=3, fig.height=5, fig.align="center"}
ggplot(data = tibble(x=-4:4))+
  aes(x = x)+
  stat_function(fun = dnorm, size=2, color="blue")+
  stat_function(fun=dnorm, xlim=c(-4,-2), geom="area", fill="blue", alpha=0.5)+
  stat_function(fun=dnorm, xlim=c(2,4), geom="area", fill="blue", alpha=0.5)+
  labs(x = "t",
       y = "Probability")+
  scale_x_continuous(breaks=c(-2,2),
                      label=c(expression(-t[i]), expression(t[i])))+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=20)
```

# Hypothesis Tests in Regression Output I

```{r, echo = T}
summary(school_reg)
```

# Hypothesis Tests in Regression Output II

- In `broom`'s `tidy()` (with confidence intervals)

```{r, echo = T}
tidy(school_reg, conf.int=TRUE)
```

# Conclusions

- .whisper[We have sufficient evidence to reject `\\(H_0\\)` in favor of our alternative hypothesis, that there *is a relationship* between class size and test scores].
    -

- .whisper[We are 95% confident that the true relationship is between `\\(-3.22\\)` and `\\(-1.34\\)`.]

# Hypothesis Testing and Confidence Intervals: Relationship

- Confidence intervals are all *two-sided* by nature
$$CI_{0.95}=\left(\left[\hat{\beta_1}-2 \times se(\hat{\beta_1})\right], \, \left[\hat{\beta_1}+2 \times se(\hat{\beta_1}\right]) \right)$$

- Hypothesis test $(t$-test) of $H_0: \, \beta_1=0$ computes a $t$-value of<sup>.red[1]</sup>
$$t=\frac{\hat{\beta_1}}{se(\hat{\beta_1})}$$
and $p$<0.05 when $t\geq2$

- .onfire[If a confidence interval contains the $H_0$ value (i.e. $0$, for our test), then *we fail to reject* $H_0$.]

.footnote[<sup>.red[1]</sup> Technically the `\\(t\\)`-value is `\\(\frac{\hat{\beta_1}-\beta_{1,0}}{se(\hat{\beta_1})}\\)`, but our `\\(H_0\\)` is often that `\\(beta_{1,0}=0\\)`, such that the formula reduces to the one above.]

# Significance In Regression Tables

.pull-left[
.font80[
.regtable[
```{r}
library(huxtable)
huxreg("Test Score" = school_reg,
       coefs = c("Intercept" = "(Intercept)",
                 "STR" = "str"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```
]
]
]

- Statistical significance is shown by asterisks, common (but not always!) standard:
    - 1 asterisk: significant at $\alpha=0.10$
    - 2 asterisks: significant at $\alpha=0.05$
    - 3 asterisks: significant at $\alpha=0.01$

- Rare, but sometimes regression tables include $p$-values for estimates
]
